---
title: "Part 1: Ecoregion-Based Range Inference"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Part 1: Ecoregion-Based Range Inference}
  %\VignetteEngine{knitr::rmarkdown_notangle}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)

library(gbif.range)

ext_file <- function(...) {
  path <- system.file("extdata", ..., package = "gbif.range")
  if (nzchar(path)) {
    return(path)
  }
  normalizePath(file.path("..", "inst", "extdata", ...), mustWork = TRUE)
}
```

## Scope

This vignette focuses on the core ecological idea behind `gbif.range`: species ranges are inferred from occurrences, but those occurrences are constrained by ecoregions rather than expanded across the landscape as a purely geometric hull.

The two longer examples cover a broad terrestrial workflow using `Panthera tigris` and a finer regional workflow using `Arctostaphylos alpinus` in the European Alps with custom ecoregions derived from environmental rasters.

The relevant functions are:

- `get_range()` for the actual range inference,
- `read_ecoreg()` and `ecoreg_list` for packaged large-scale ecoregions,
- `make_ecoreg()` for custom ecoregions derived from environmental rasters,
- `cv_range()` and `evaluate_range()` for evaluating the resulting maps.

## How `get_range()` works

`get_range()` is designed for one focal species at a time. The algorithm can be summarized as four linked operations:

1. remove isolated spatial outliers from the occurrence table,
2. identify occupied ecoregions and cluster occurrences within them,
3. build convex hulls and buffers for each within-ecoregion cluster,
4. intersect those polygons with the occupied ecoregions to enforce ecological limits.

This means that the final output reflects both the geometry of the occurrences and the eco-geographic structure of the study area.

## A minimal offline example

The example below is deliberately small and fully offline. Two synthetic rasters are used to create a custom ecoregion layer, then a short occurrence table is passed to `get_range()`.

```{r minimal-example}
set.seed(1)

# Create two simple environmental surfaces on a small study area.
r1 <- terra::rast(ncols = 20, nrows = 20, xmin = 0, xmax = 10, ymin = 0, ymax = 10)
terra::values(r1) <- rep(seq(0, 1, length.out = 20), each = 20)
r2 <- terra::rast(r1)
terra::values(r2) <- rep(seq(0, 1, length.out = 20), times = 20)
env <- c(r1, r2)

# Derive four custom ecoregions from the environmental layers.
eco <- make_ecoreg(env = env, nclass = 4)

# Define one species as an occurrence table in the format expected by get_range().
occ <- data.frame(
  input_search = "example_species",
  decimalLongitude = c(0.5, 1.2, 2.4, 3.6, 2.8, 6.0, 7.2, 7.4, 7.6, 8.8),
  decimalLatitude = c(1.0, 0.1, 2.3, 2.5, 2.7, 5.0, 7.1, 7.3, 6.5, 7.7)
)

# Build the range from the occurrences and the ecoregion layer.
range_obj <- get_range(
  occ_coord = occ,
  ecoreg = eco,
  ecoreg_name = "EcoRegion",
  verbose = FALSE
)

terra::plot(
  range_obj$rangeOutput,
  col = "#3c8d5a",
  main = "A minimal ecoregion-constrained range"
)
points(occ$decimalLongitude, occ$decimalLatitude, pch = 4)
```

Three practical points are worth noting here.

First, the occurrence object must contain decimal longitude and latitude. In practice it will often be a `getGBIF` object returned by `get_gbif()`, but a regular `data.frame` is also valid.

Second, custom ecoregions created with `make_ecoreg()` can be passed directly into `get_range()`. This is useful for regional studies where global ecoregions are too coarse.

Third, `get_range()` returns a `getRange` object. The spatial output is stored in `range_obj$rangeOutput`, while the original inputs are kept in `range_obj$init.args`.

## Packaged versus custom ecoregions

Large-scale analyses usually rely on one of the packaged ecoregion layers:

```{r ecoregions}
vapply(ecoreg_list, `[[`, character(1), "filename")
```

The public pattern for a terrestrial analysis is:

```{r built-in-ecoregions, eval = FALSE}
eco_terra <- read_ecoreg("eco_terra")

range_tiger <- get_range(
  occ_coord = obs_tiger,
  ecoreg = eco_terra,
  ecoreg_name = "ECO_NAME"
)
```

In contrast, regional analyses often benefit from custom ecoregions. The package ships with raster examples under `inst/extdata`, including `rst.tif`, which can be used as a starting point for a fine-grained environmental classification:

```{r custom-ecoregions, eval = FALSE}
rst <- terra::rast(ext_file("rst.tif"))
my_eco <- make_ecoreg(env = rst, nclass = 200)

range_regional <- get_range(
  occ_coord = obs_regional,
  ecoreg = my_eco,
  ecoreg_name = "EcoRegion",
  res = 0.05
)
```

The key trade-off is ecological detail versus robustness to sampling density. Coarser ecoregions produce broader, smoother ranges. Finer ecoregions can better capture regional structure, but they also make the results more sensitive to local sampling gaps.

## A broad terrestrial workflow: *Panthera tigris*

Tiger occurrences are a good example of the broad terrestrial use case for `gbif.range`. The workflow is straightforward: retrieve occurrences with a coarse precision filter, load the terrestrial ecoregions of the world, and infer the range inside those ecoregional boundaries.

```{r tiger-workflow, eval = FALSE}
# Step 1: download global tiger occurrences with a 100 km precision filter.
obs_tiger <- get_gbif(
  sp_name = "Panthera tigris",
  grain = 100
)

# Step 2: load the packaged terrestrial ecoregion layer.
eco_terra <- read_ecoreg("eco_terra")

# Step 3: infer the range at the default 0.1 degree output resolution.
range_tiger <- get_range(
  occ_coord = obs_tiger,
  ecoreg = eco_terra,
  ecoreg_name = "ECO_NAME",
  degrees_outlier = 5,
  clust_pts_outlier = 4,
  res = 0.1
)

terra::plot(range_tiger$rangeOutput, col = "#238b45")
points(obs_tiger[, c("decimalLongitude", "decimalLatitude")], pch = 20)
```

This is the typical continental to global pattern. The ecoregion layer is externally defined and ecologically interpretable, while the occurrence filter is intentionally conservative. In practice, this is the kind of workflow where `gbif.range` is most useful as a transparent alternative to unconstrained geometric envelopes.

## A finer regional workflow: *Arctostaphylos alpinus* in the Alps

The second workflow is deliberately different. Here the issue is not worldwide coverage, but the opposite: the global terrestrial ecoregions are too coarse for a mountain system with strong local climatic structure. A custom ecoregion layer is therefore built from packaged example rasters.

```{r arcto-workflow, eval = FALSE}
# Step 1: define the Alps study region and retrieve occurrences.
alps_extent <- terra::vect(ext_file("shp_lonlat.shp"))

obs_arcto <- get_gbif(
  sp_name = "Arctostaphylos alpinus",
  geo = alps_extent,
  grain = 1
)

# Step 2: create finer ecoregions from the packaged environmental rasters.
rst <- terra::rast(ext_file("rst.tif"))
eco_alps <- make_ecoreg(env = rst, nclass = 200)

# Step 3: infer the range at finer output resolution.
range_arcto <- get_range(
  occ_coord = obs_arcto,
  ecoreg = eco_alps,
  ecoreg_name = "EcoRegion",
  res = 0.05
)

terra::plot(range_arcto$rangeOutput, col = "#3c8d5a")
points(obs_arcto[, c("decimalLongitude", "decimalLatitude")], pch = 20)
```

This regional example shows why `make_ecoreg()` is part of the main package workflow rather than an auxiliary convenience. At small spatial extents, the ecological realism of the range map often depends more on the choice of ecoregion layer than on minor changes in buffer parameters.

## Tuning the main range arguments

The most important `get_range()` arguments usually fall into three groups.

`degrees_outlier` and `clust_pts_outlier` control how aggressively isolated occurrences are removed. These matter most when the input contains obvious anomalies or disjunct stray clusters.

`buff_width_point`, `buff_incrmt_pts_line`, and `buff_width_polygon` control how singletons, linear clusters, and polygon hulls are buffered before the final ecoregion intersection.

`raster`, `format`, and `res` control the output type. In many exploratory workflows a vector output is convenient, whereas gridded outputs are useful when later stacking many species maps.

A useful rule of thumb is this: at broad scale, the defaults for outlier detection are often a reasonable starting point because the main goal is to remove very isolated records or obvious anomalies. At finer regional scale, the more consequential decisions are often the observation precision filter used upstream in `get_gbif()` and the granularity of the ecoregion layer generated by `make_ecoreg()`.

## Evaluation workflows

The package provides two complementary ways to evaluate range maps.

`cv_range()` performs internal cross-validation of a `getRange` object by repeatedly rebuilding the map from subsets of the original occurrences:

```{r cv-code, eval = FALSE}
cv_res <- cv_range(
  range_object = range_obj,
  cv = "block-cv",
  nfolds = 3,
  nblocks = 2,
  backpoints = 500
)
```

`evaluate_range()` compares saved range outputs with external validation layers. The package includes a small validation example under `inst/extdata`:

```{r evaluate-code, eval = FALSE}
root_dir <- system.file("extdata", package = "gbif.range")

res_eval <- evaluate_range(
  root_dir = root_dir,
  valData_dir = "SDM",
  ecoRM_dir = "EcoRM",
  print_map = FALSE,
  verbose = FALSE
)

head(res_eval$df_eval)
```

Together, `cv_range()` and `evaluate_range()` let you tune parameter choices on well-documented taxa before applying the same workflow to data-poor species.

The same logic can be used with the broad and regional examples above. For example, a block cross-validation of the tiger workflow is:

```{r tiger-cv, eval = FALSE}
cv_range(
  range_object = range_tiger,
  cv = "block-cv",
  nfolds = 5,
  nblocks = 2,
  backpoints = 1e4
)
```

That kind of evaluation is particularly useful when deciding how strongly to filter outliers, how fine the ecoregion layer should be, or whether a raster output resolution is appropriate for later biodiversity summaries.

## Take-home message

`get_range()` is best understood as an ecological range-construction tool rather than a pure hull algorithm. The strength of the approach lies in combining occurrence geometry with eco-geographic structure. The rest of the package, including GBIF retrieval and large disk-based workflows, is built to feed that core mapping step in a transparent and reproducible way.
