---
title: "Part 2: GBIF Retrieval, Taxonomy, and Filtering"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Part 2: GBIF Retrieval, Taxonomy, and Filtering}
  %\VignetteEngine{knitr::rmarkdown_notangle}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)

library(gbif.range)

ext_file <- function(...) {
  path <- system.file("extdata", ..., package = "gbif.range")
  if (nzchar(path)) {
    return(path)
  }
  normalizePath(file.path("..", "inst", "extdata", ...), mustWork = TRUE)
}
```

## Scope

This vignette covers the package components that sit upstream of range inference:

- `get_status()` to inspect the GBIF backbone taxon concept and synonym set,
- `get_gbif_count()` to gauge likely download size,
- `get_gbif()` to retrieve occurrences without GBIF credentials,
- `obs_filter()` to thin or aggregate occurrences for downstream work,
- `make_tiles()` to build explicit GBIF geometry tiles,
- `get_doi()` to turn one or more downloads into a citable GBIF-derived DOI.

## Why taxonomy comes first

In `gbif.range`, taxonomy is not an afterthought. The package uses the GBIF backbone to decide which accepted taxon concept a query refers to and which synonyms belong to that concept.

This is why `get_status()` is often the best first step:

```{r get-status, eval = FALSE}
tax <- get_status("Cypripedium calceolus", all = FALSE)
tax
```

When `all = FALSE`, the output is deliberately close to the logic used by `get_gbif()`: it returns the accepted name and the synonyms that are actually used for occurrence retrieval. When `all = TRUE`, additional related and child names are included for taxonomic inspection, but those extra names are not used to download occurrences.

The practical consequence is important: `get_gbif()` harmonizes the query to the accepted GBIF taxon key, but the returned occurrence table still preserves both record-level and accepted-name fields.

## Count before download

For broad extents or common taxa, a quick count helps you decide whether a direct retrieval is appropriate:

```{r get-gbif-count, eval = FALSE}
get_gbif_count("Panthera tigris", search = TRUE)
```

This is particularly useful if you need to decide between:

- a direct `get_gbif()` call,
- a more strongly filtered call with a stricter `grain` or `occ_samp`,
- or a downloaded GBIF export that will later be processed with the disk-based batch workflow described in Part 3.

## Credential-free GBIF retrieval

`get_gbif()` is a credential-free wrapper around `rgbif::occ_search()`. The main package contribution is that it combines taxonomic harmonization, geographic tiling of large extents, and a practical sequence of post-download filters in one function.

```{r get-gbif, eval = FALSE}
# Download global tiger occurrences with the default filters.
obs_tiger <- get_gbif("Panthera tigris", grain = 100)

# Inspect the accepted name and synonym mapping used internally.
get_status("Panthera tigris", all = FALSE)

# Retrieve a terrestrial ecoregion layer and build the range.
eco_terra <- read_ecoreg("eco_terra")
range_tiger <- get_range(
  occ_coord = obs_tiger,
  ecoreg = eco_terra,
  ecoreg_name = "ECO_NAME"
)
```

The arguments most often worth adjusting are:

- `grain`, which filters by coordinate uncertainty and decimal precision,
- `basis` and `establishment`, which control the allowed record types,
- `time_period`, if the analysis targets a defined time window,
- `occ_samp`, if the geographic extent is so large that a full retrieval is unnecessary or impractical.

## Post-download thinning with `obs_filter()`

The package also includes a lightweight grid-based thinning helper. This is useful when many records fall into the same grid cell and you want one retained observation per species per cell before a downstream analysis.

The example below uses the bundled offline GBIF-style example table rather than a live web query.

```{r obs-filter}
occ_raw <- utils::read.delim(ext_file("occ_example_4sps.csv"), sep = "\t", stringsAsFactors = FALSE)
occ_raw$input_search <- occ_raw$species
occ_gbif <- gbif.range:::getGBIF(occ_raw)

# Build a coarse grid that spans the example records.
grid <- terra::rast(
  xmin = min(occ_gbif$decimalLongitude) - 1,
  xmax = max(occ_gbif$decimalLongitude) + 1,
  ymin = min(occ_gbif$decimalLatitude) - 1,
  ymax = max(occ_gbif$decimalLatitude) + 1,
  resolution = 10,
  crs = "EPSG:4326"
)

# Keep at most one record per species and grid cell.
obs_thin <- obs_filter(occ_gbif, grid)
head(obs_thin)
```

This kind of thinning does not replace ecological cleaning or taxonomic checks, but it can be a useful way to reduce local clustering before plotting or model calibration.

## Explicit tiling workflows

`get_gbif()` handles tiling internally, but `make_tiles()` is available when you want to create GBIF-ready geometry tiles yourself:

```{r make-tiles, eval = FALSE}
tiles <- make_tiles(terra::ext(-20, 40, 0, 60), nsize = 5)
tiles[[1]]
```

This is mostly useful for custom `rgbif` workflows or explicit diagnostics of how an extent is being subdivided.

## Reproducible GBIF citation

Once one or more `get_gbif()` calls have been made, `get_doi()` can be used to generate a GBIF-derived DOI reference for those records:

```{r get-doi, eval = FALSE}
doi_url <- get_doi(obs_tiger)
doi_url
```

This is a small function, but it is scientifically important because it improves traceability and citation of the exact GBIF-derived datasets used in an analysis.

## Take-home message

The GBIF-facing side of `gbif.range` is built around a clear sequence:

1. inspect the GBIF backbone taxon concept with `get_status()`,
2. count likely record volume with `get_gbif_count()`,
3. download filtered occurrences with `get_gbif()`,
4. optionally thin or aggregate them with `obs_filter()`,
5. cite the resulting GBIF-derived datasets with `get_doi()`.

That sequence keeps taxonomic interpretation, record selection, and reproducibility visible before range inference begins.
