---
title: "Ecologically Informed Species Range Mapping with gbif.range"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ecologically Informed Species Range Mapping with gbif.range}
  %\VignetteEngine{knitr::rmarkdown_notangle}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)

library(gbif.range)

has_data_table <- requireNamespace("data.table", quietly = TRUE)
ext_file <- function(...) {
  path <- system.file("extdata", ..., package = "gbif.range")
  if (nzchar(path)) {
    return(path)
  }
  normalizePath(file.path("..", "inst", "extdata", ...), mustWork = TRUE)
}
```

## Overview

`gbif.range` is built around a simple scientific idea: species ranges can be approximated from occurrence records when those records are constrained by ecologically meaningful spatial units. In the package, those units are ecoregions supplied either from bundled large-scale layers or from custom classifications created with environmental rasters.

The package covers three main workflow families:

1. ecologically constrained range inference through ecoregion-constrained convex-buffer hulls,
2. credential-free, synonym-aware retrieval of GBIF records over large extents, and
3. disk-based batch processing for very large GBIF tables already downloaded to disk.

These workflows are complemented by tools to inspect GBIF taxonomy, thin or aggregate occurrence data, generate custom ecoregions, evaluate range outputs, and create GBIF-derived DOIs.

## Main Functions

```{r function-table, echo = FALSE}
fun_tab <- data.frame(
  Function = c(
    "get_range()",
    "read_ecoreg() / get_ecoreg()",
    "make_ecoreg()",
    "get_status()",
    "get_gbif_count() / get_gbif()",
    "obs_filter()",
    "split_gbif_by_species()",
    "species_csvs_to_ranges()",
    "evaluate_range() / cv_range()",
    "get_doi() / make_tiles()"
  ),
  Role = c(
    "Infer species ranges from occurrences and ecoregions.",
    "Load packaged ecoregion layers or download them if needed.",
    "Build custom ecoregion layers from environmental rasters.",
    "Inspect the GBIF backbone taxon concept and synonym set.",
    "Count or download GBIF occurrences without credentials.",
    "Thin or aggregate occurrences on a user-defined grid.",
    "Split a large GBIF table into one file per species.",
    "Read per-species files and build one range output per species.",
    "Evaluate inferred ranges against external data or by cross-validation.",
    "Support reproducible GBIF citation and explicit GBIF tiling workflows."
  ),
  stringsAsFactors = FALSE
)

knitr::kable(fun_tab, align = c("l", "l"))
```

## 1. Ecologically Constrained Range Inference

The central function, `get_range()`, combines occurrence data and ecoregions in four broad steps:

1. occurrence points are filtered for isolated spatial outliers,
2. remaining points are clustered within occupied ecoregions,
3. cluster-specific convex hulls and buffers are built, and
4. those geometries are intersected with their parent ecoregions to keep the final range ecologically constrained.

This is the main distinction from a plain geometric hull: the final map is not allowed to spill across ecoregion boundaries that were not occupied by the focal species.

### Minimal offline example

The example below uses two synthetic environmental rasters to create a small custom ecoregion map with `make_ecoreg()`, then infers a range from an in-memory occurrence table.

```{r minimal-range}
set.seed(1)

r1 <- terra::rast(ncols = 20, nrows = 20, xmin = 0, xmax = 10, ymin = 0, ymax = 10)
terra::values(r1) <- rep(seq(0, 1, length.out = 20), each = 20)
r2 <- terra::rast(r1)
terra::values(r2) <- rep(seq(0, 1, length.out = 20), times = 20)
env <- c(r1, r2)

eco <- make_ecoreg(env = env, nclass = 4)

occ <- data.frame(
  input_search = "example_species",
  decimalLongitude = c(0.5, 1.2, 2.4, 3.6, 2.8, 6.0, 7.2, 7.4, 7.6, 8.8),
  decimalLatitude = c(1.0, 0.1, 2.3, 2.5, 2.7, 5.0, 7.1, 7.3, 6.5, 7.7)
)

range_obj <- get_range(
  occ_coord = occ,
  ecoreg = eco,
  ecoreg_name = "EcoRegion",
  verbose = FALSE
)

terra::plot(range_obj$rangeOutput, col = "#3c8d5a",
  main = "Range inferred from custom ecoregions")
points(occ$decimalLongitude, occ$decimalLatitude, pch = 4)
```

This example is intentionally simple, but it shows the main logic of the package: a species range is inferred from the occupied eco-geographic units rather than from coordinates alone.

## 2. Working with Ecoregions

`gbif.range` accepts either:

- a bundled large-scale ecoregion layer loaded with `read_ecoreg()`, or
- a custom layer created with `make_ecoreg()`.

The built-in layers currently include terrestrial, freshwater, and marine products:

```{r ecoreg-list}
vapply(ecoreg_list, `[[`, character(1), "filename")
```

For broad terrestrial applications, the standard workflow is:

```{r built-in-ecoreg, eval = FALSE}
eco_terra <- read_ecoreg("eco_terra")

range_tiger <- get_range(
  occ_coord = obs_tiger,
  ecoreg = eco_terra,
  ecoreg_name = "ECO_NAME"
)
```

For finer regional work, a custom ecoregion layer is often preferable because coarse global ecoregions may be too broad for local environmental gradients.

## 3. GBIF Retrieval, Taxonomy, and Filtering

### Synonym-aware taxonomic inspection

Before downloading occurrences, `get_status()` can be used to inspect how GBIF resolves a submitted name. This is especially useful when synonymy or taxonomic ambiguity matters.

```{r status-code, eval = FALSE}
tax <- get_status("Panthera tigris", all = FALSE)
tax
```

When `all = FALSE`, the returned rows represent the accepted GBIF taxon concept and the synonyms that `get_gbif()` will use internally. In other words, the query is harmonized to the accepted GBIF taxon key, while the resulting occurrence table still retains both record-level and accepted-name fields.

### Credential-free GBIF retrieval over large extents

`get_gbif()` is a lightweight wrapper around `rgbif::occ_search()`. For large extents it tiles the request automatically, which makes it useful when you want a fully programmatic workflow without GBIF credentials.

```{r gbif-code, eval = FALSE}
# Count first if the expected volume is uncertain.
get_gbif_count("Panthera tigris", search = TRUE)

# Download global tiger occurrences with default filtering.
obs_tiger <- get_gbif("Panthera tigris", grain = 100)

# Build a range from the accepted name plus its synonyms.
eco_terra <- read_ecoreg("eco_terra")
range_tiger <- get_range(
  occ_coord = obs_tiger,
  ecoreg = eco_terra,
  ecoreg_name = "ECO_NAME"
)
```

If the expected number of records is extremely large, `occ_samp` can be used to subsample within tiles. For explicit geometry workflows outside `get_gbif()`, `make_tiles()` can generate GBIF-ready polygons:

```{r make-tiles-code, eval = FALSE}
tiles <- make_tiles(terra::ext(-20, 40, 0, 60), nsize = 5)
tiles[[1]]
```

### Post-download thinning and aggregation

`obs_filter()` works on `getGBIF` objects and can reduce multiple records per grid cell to a single retained occurrence per species. The offline example below uses the bundled two-species occurrence table.

```{r obs-filter}
occ_raw <- utils::read.delim(ext_file("occ_example_4sps.csv"), sep = "\t", stringsAsFactors = FALSE)
occ_raw$input_search <- occ_raw$species
occ_gbif <- gbif.range:::getGBIF(occ_raw)

grid <- terra::rast(
  xmin = min(occ_gbif$decimalLongitude) - 1,
  xmax = max(occ_gbif$decimalLongitude) + 1,
  ymin = min(occ_gbif$decimalLatitude) - 1,
  ymax = max(occ_gbif$decimalLatitude) + 1,
  resolution = 10,
  crs = "EPSG:4326"
)

obs_thin <- obs_filter(occ_gbif, grid)
head(obs_thin)
```

### Reproducible GBIF citation

Once one or more `get_gbif()` outputs have been created, `get_doi()` can be used to produce a derived GBIF DOI:

```{r doi-code, eval = FALSE}
doi_url <- get_doi(obs_tiger)
doi_url
```

## 4. Evaluation and Validation

`gbif.range` includes two complementary evaluation paths:

- `evaluate_range()` compares generated range maps with external validation products such as rasters or polygons.
- `cv_range()` rebuilds a `getRange` object repeatedly from subsets of its own occurrences, using either random or spatial block cross-validation.

The package ships with a small external validation example under `inst/extdata`:

```{r evaluate-code, eval = FALSE}
root_dir <- system.file("extdata", package = "gbif.range")

res_eval <- evaluate_range(
  root_dir = root_dir,
  valData_dir = "SDM",
  ecoRM_dir = "EcoRM",
  print_map = FALSE,
  verbose = FALSE
)

head(res_eval$df_eval)
```

A lightweight cross-validation call follows the same logic:

```{r cv-code, eval = FALSE}
cv_res <- cv_range(
  range_object = range_obj,
  cv = "block-cv",
  nfolds = 3,
  nblocks = 2,
  backpoints = 500
)
```

These tools are particularly useful for calibrating ecoregion choice, outlier thresholds, and buffering settings before scaling the workflow to many species.

## 5. Scalable Batch Processing from Downloaded GBIF Tables

When occurrence data have already been downloaded from GBIF as a large multi-species table, the package now supports a disk-based workflow that avoids reading the full table into memory.

The batch workflow is deliberately split into two steps:

1. `split_gbif_by_species()` writes one species file per GBIF taxon key.
2. `species_csvs_to_ranges()` reads those files sequentially and applies `get_range()` one species at a time.

This design matches the scientific assumptions of `get_range()`, which is intended to operate on one species per call.

### Step 1: split a downloaded GBIF table

```{r split-batch, eval = has_data_table}
gbif_file <- ext_file("occ_example_4sps.csv")

batch_root <- file.path(tempdir(), "gbif-range-vignette")
split_dir <- file.path(batch_root, "split")
occ_dir <- file.path(batch_root, "occ_min")
range_dir <- file.path(batch_root, "ranges")

unlink(batch_root, recursive = TRUE)

split_summary <- split_gbif_by_species(
  input_file = gbif_file,
  outdir = split_dir,
  chunk_size = 100,
  sep_in = "\t",
  sep_out = "\t",
  overwrite = TRUE,
  verbose = FALSE
)

knitr::kable(split_summary[, c("species_name", "n_records", "species_file")], align = "l")
```

The output files are still tab-delimited by default, but they are saved with a `.csv` extension for convenience and follow the pattern `occurrences_speciesKey_<key>_<species>.csv`.

### Step 2: build one range per species from disk

For the vignette we use one simple enclosing polygon as the ecoregion layer so the example stays fully offline and reproducible during package checks. In ordinary use, the same workflow can be run directly with `ecoreg = "eco_terra"` or with a spatial object returned by `read_ecoreg("eco_terra")`.

```{r ranges-from-disk, eval = has_data_table, fig.width = 7, fig.height = 5}
occ_example <- utils::read.delim(gbif_file, sep = "\t", stringsAsFactors = FALSE)

eco_batch <- sf::st_sf(
  data.frame(ECO_NAME = "batch_demo"),
  geometry = sf::st_sfc(
    sf::st_polygon(list(matrix(
      c(
        min(occ_example$decimalLongitude) - 5, min(occ_example$decimalLatitude) - 5,
        max(occ_example$decimalLongitude) + 5, min(occ_example$decimalLatitude) - 5,
        max(occ_example$decimalLongitude) + 5, max(occ_example$decimalLatitude) + 5,
        min(occ_example$decimalLongitude) - 5, max(occ_example$decimalLatitude) + 5,
        min(occ_example$decimalLongitude) - 5, min(occ_example$decimalLatitude) - 5
      ),
      ncol = 2,
      byrow = TRUE
    ))),
    crs = 4326
  )
)


range_summary <- species_csvs_to_ranges(
  species_dir = split_dir,
  ecoreg = eco_batch,
  ecoreg_name = "ECO_NAME",
  outdir = range_dir,
  occ_outdir = occ_dir,
  occ_save_as = "tsv",
  range_save_as = "rds",
  sep_in = "\t",
  overwrite = TRUE,
  degrees_outlier = 30,
  clust_pts_outlier = 2,
  buff_width_point = 1,
  buff_incrmt_pts_line = 0.1,
  buff_width_polygon = 1,
  raster = FALSE,
  format = "SpatVector",
  verbose = FALSE
)

knitr::kable(range_summary[, c("species_name", "n_points", "range_file")], align = "l")
```

The same call can also resolve packaged ecoregions internally:

```{r ranges-from-disk-shortcut, eval = FALSE}
range_summary <- species_csvs_to_ranges(
  species_dir = split_dir,
  ecoreg = "eco_terra",
  ecoreg_name = "ECO_NAME",
  outdir = range_dir,
  overwrite = TRUE
)
```

### Step 3: read saved ranges back from disk

The helper `read_range_rds()` reads the saved `.rds` range files back into a regular list containing `init.args` and `rangeOutput`.

```{r read-range-rds, eval = has_data_table}
# read one range file to inspect the output structure
range_one <- read_range_rds(range_summary$range_file[1])

# plot all ranges in the summary
#PLOT EMPTY PLOT FIRST
terra::plot(range_one$rangeOutput, main = "All batch-generated ranges", col = rgb(0.1, 0.1, 0.1, 0.2))
# plot empty, just based on extent, no col
plot(terra::ext(range_one$rangeOutput), main = "All batch-generated ranges", legend=FALSE)
transp <- 0.2
color_vec <- c(rgb(0.1, 0.1, 0.1, transp), rgb(0.8, 0.3, 0.1, transp), rgb(0.1, 0.3, 0.8, transp), rgb(0.8, 0.8, 0.1, transp))
for (range_file in range_summary$range_file) {
  
  range_obj <- read_range_rds(range_file)
  terra::plot(
    range_obj$rangeOutput,
    add = TRUE,
    col = rgb(0.1, 0.1, 0.1, transp)# color_vec[which(range_summary$range_file == range_file)]
  )
}
# plot map land contour
maps::map("world", bg = "lightblue", add=T)
# maps::map("world", add = TRUE)
```

In practical use, this disk-based workflow is the most scalable option for large downloaded GBIF tables or cluster-style batch processing where species are processed one at a time.

## Closing Remarks

The package is most useful when its components are combined deliberately:

- use `get_status()` and `get_gbif_count()` to understand the taxon concept and expected record volume,
- use `get_gbif()` or a downloaded GBIF table as the occurrence source,
- choose an ecoregion layer that matches the spatial grain and ecological question,
- build ranges with `get_range()` or the disk-based batch workflow, and
- evaluate decisions with `evaluate_range()` or `cv_range()` whenever independent data or internal tuning are possible.

That combination makes `gbif.range` more than a single hull function: it is a complete workflow for building ecologically constrained, transparent, and reproducible range maps from occurrence data.
