\documentclass[a4paper]{book}
\usepackage[times,inconsolata,hyper]{Rd}
\usepackage{makeidx}
\usepackage[utf8]{inputenc} % @SET ENCODING@
% \usepackage{graphicx} % @USE GRAPHICX@
\makeindex{}
\begin{document}
\chapter*{}
\begin{center}
{\textbf{\huge wsl.gbif}}
\par\bigskip{\large \today}
\end{center}
\inputencoding{utf8}
\ifthenelse{\boolean{Rd@use@hyper}}{\hypersetup{pdftitle = {wsl.gbif: A toolbox to efficiently download and filter large GBIF observational datasets for sound spatial analyses}}}{}\begin{description}
\raggedright{}
\item[Type]\AsIs{Package}
\item[Title]\AsIs{A toolbox to efficiently download and filter large GBIF observational datasets for sound spatial analyses}
\item[Version]\AsIs{0.1.0}
\item[Depends]\AsIs{R (>= 4.0.0), raster, terra,  rgbif, CoordinateCleaner}
\item[Description]\AsIs{Package aiming at easing the workflow of retrieving GBIF observations at large spatial scale for all species accepted names and synonyms, and filtering them according to the specific scale of an analysis}
\item[License]\AsIs{GPL (>=3)}
\item[BugReports]\AsIs{}\url{https://github.com/8Ginette8/wsl.gbif/issues}\AsIs{}
\item[Encoding]\AsIs{UTF-8}
\item[Maintainer]\AsIs{Yohann Chauvier }\email{yohann.chauvier@wsl.ch}\AsIs{}
\item[LazyData]\AsIs{true}
\item[RoxygenNote]\AsIs{7.1.2}
\item[Authors]\AsIs{Yohann Chauvier [cre,aut] (https://orcid.org/0000-0001-9399-3192)}
\item[Collate]\AsIs{'make_tiles.R'
'wsl_gbif.R'
'wsl_taXnames.R'
'wsl_doi.R'
'wsl_obs_filter.R'}
\end{description}
\Rdcontents{\R{} topics documented:}
\inputencoding{utf8}
\HeaderA{make\_tiles}{Create a specific number of tiles based on a raster extent}{make.Rul.tiles}
%
\begin{Description}\relax
Based on a specific extent, one or several tiles are generated. Tiles can be smaller
numerical extents or geometry arguments POLYGON(). The original extent is therefore either
converted into a POLYGON() argument, or divided into Ntiles of regular fragments which are
converted into POLYGON() arguments and smaller numerical extents.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
make_tiles(geo, Ntiles, sext = TRUE)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{geo}] Object of class 'Extent', 'SpatExtent', 'SpatialPolygon', 'SpatialPolygonDataframe',
or 'SpaVector' (WGS84 or planar) to define the study's area extent. Default is NULL i.e. the
whole globe.

\item[\code{Ntiles}] Numeric. In how many tiles/fragments should geo be divided approximately?

\item[\code{sext}] Logical. Should a list of numerical extents 'c(xmin,xmax,ymin,ymax)' also be returned
for each generated POLYGON()?
\end{ldescription}
\end{Arguments}
%
\begin{Value}
A list of geometry arguments POLYGON() of length Ntiles (and of numerical extents
if sext=TRUE)
\end{Value}
%
\begin{Author}\relax
Yohann Chauvier
\end{Author}
%
\begin{References}\relax
Chauvier, Y., Thuiller, W., Brun, P., Lavergne, S., Descombes, P., Karger, D. N., ... \& Zimmermann,
N. E. (2021). Influence of climate, soil, and land cover on plant species distribution in the
European Alps. Ecological monographs, 91(2), e01433. 10.1002/ecm.1433
\end{References}
%
\begin{Examples}
\begin{ExampleCode}

# Load the European Alps Extent
data(geo_dat)

# Apply the function to divide the extent in ~20 fragments
make_tiles(geo=shp.lonlat,Ntiles=20,sext=TRUE)

\end{ExampleCode}
\end{Examples}
\inputencoding{utf8}
\HeaderA{wsl\_doi}{Get a custom DOI for a GBIF filtered dataset}{wsl.Rul.doi}
%
\begin{Description}\relax
A small user friendly wrapper of the derived\_dataset() function of the
rgbif R package, compatible with one or several wsl\_gbif() outputs.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
wsl_doi(
  wsl_gbif = list(),
  title = NULL,
  description = NULL,
  source_url = "https://example.com/",
  usr = "",
  pwd = "",
  ...
)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{wsl\_gbif}] List. List of one or several wsl\_gbif outputs.

\item[\code{title}] The title for your derived dataset.

\item[\code{source\_url}] A link to where the dataset is stored.

\item[\code{pwd}] Your GBIF password.

\item[\code{...}] Additonnal parameters for derived\_dataset() in rgbif.
R package.

\item[\code{descritpion}] A description of the dataset.

\item[\code{user}] Your GBIF username.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
see derived\_dataset() function from the rgbif R package
\end{Details}
%
\begin{Value}
One citable DOI and its information.
\end{Value}
%
\begin{References}\relax
Chamberlain, S., Oldoni, D., \& Waller, J. (2022). rgbif: interface to the global biodiversity
information facility API. 10.5281/zenodo.6023735
\end{References}
%
\begin{SeeAlso}\relax
The rgbif package for additional and more general approaches to get GBIF DOI
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}

# Downloading worldwide the observations of Panthera tigris and Ailuropoda melanoleuca
test1 = wsl_gbif("Panthera tigris")
test32 = wsl_gbif("Ailuropoda melanoleuca")

# Just an example on how to retrieve the DOI for only one dataset


# Just an example on how to retrieve the DOI for only one dataset


d.target = table(test1$datasetKey)
d.summary = data.frame(datasetKey = names(d.target),count = as.numeric(d.target))
rgbif::derived_dataset(d.summary,"GBIF_test",
    "Filetred and cleaned based on CoordinateCleaner",source_url="https://example.com/",
    user="your_gbif_user",pwd="your_gbif_password")

\end{ExampleCode}
\end{Examples}
\inputencoding{utf8}
\HeaderA{wsl\_gbif}{Massively download and filter GBIF observations for sound spatial analyses}{wsl.Rul.gbif}
%
\begin{Description}\relax
Implement an user-friendly workflow to download and clean gbif taxa observations.
The function uses the rgbif R package but (1) implements the same search result 
found if www.gbif.org is employed i.e., based on the input taxa name, all species
records related to its accepted name and synonyms are extracted. The function
also (2) bypasses rgbif hard limit on the number of records (100'000 max).
For this purpose, a dynamic moving window is created and used across the geographic
extent defined by the user. This window automatically fragments the specified
study area in succesive tiles of different sizes, until all tiles include < 100'000
observations. The function also (3) automatically applies a post-filtering of
observations based on the chosen resolution of the study/analysis and by partly
employing the CoordinateCleaner R package. Filtering options may be chosen and
involve several choices: study's extent, removal of duplicates, removal of absences,
basis of records selection, removal of invalid/uncertain xy coordinates (WGS84), time
period selection and removal of raster centroids. By default, the argument
hasGeospatialIssue in occ\_search() (implemented rgbif function) is set to FALSE.
To get the custom DOI of the downloaded GBIF data, the derived\_dataset() function
from the rgbif package must be used with the column 'datasetKey' of one or several
outputs.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
wsl_gbif(
  sp_name = NULL,
  conf_match = 90,
  geo = NULL,
  grain = 1000,
  duplicates = FALSE,
  absences = FALSE,
  no_xy = FALSE,
  basis = c("OBSERVATION", "HUMAN_OBSERVATION", "MACHINE_OBSERVATION",
    "MATERIAL_SAMPLE", "PRESERVED_SPECIMEN", "FOSSIL_SPECIMEN", "LIVING_SPECIMEN",
    "LITERATURE", "UNKNOWN"),
  add_infos = NULL,
  time_period = c(1000, 3000),
  identic_xy = FALSE,
  wConverted_xy = FALSE,
  centroids = FALSE,
  ntries = 10,
  error.skip = TRUE,
  ...
)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{sp\_name}] Character. Scientific name to run an online search
(i.e. with GBIF-API) for species observations. Works also for genus and higher taxa
levels.

\item[\code{conf\_match}] Numeric from 0 to 100. Determine the confidence threshold of match
of 'sp\_name' with the GBIF backbone taxonomy. Default is 90.

\item[\code{geo}] Object of class 'Extent', 'SpatExtent', 'SpatialPolygon', 'SpatialPolygonDataframe',
or 'SpaVector' (WGS84) to define the study's area extent. Default is NULL i.e. the whole globe.

\item[\code{grain}] Numeric. Specify in meters the study resolution. Used to
filter gbif records (x2) according to their uncertainties and number of coordinate
decimals. Records with no information on coordinate uncertainties (column
'coordinateUncertaintyInMeters') are be kept by default. See details.

\item[\code{duplicates}] Logical. Should duplicated records be kept?

\item[\code{absences}] Logical. Should absence records be kept?

\item[\code{no\_xy}] Logical. Default is FALSE i.e. only records with coordinates are
downloaded. If TRUE, only records with no coordinates are downloaded.

\item[\code{basis}] Character. Which basis of records should be selected?
Default is all i.e. c("OBSERVATION", "HUMAN\_OBSERVATION", "MACHINE\_OBSERVATION",
"MATERIAL\_SAMPLE", "PRESERVED\_SPECIMEN", "FOSSIL\_SPECIMEN", "LIVING\_SPECIMEN", "LITERATURE",
"UNKNOWN"). Description may be found here: https://docs.gbif.org/course-data-use/en/basis-of-record.html

\item[\code{add\_infos}] Character. Infos that may be added to the default output information.
List of IDs may be found at: https://www.gbif.org/developer/occurrence.
Default IDs contain 'taxonKey', 'scientificName', 'acceptedTaxonKey',
'acceptedScientificName', 'individualCount', 'decimalLatitude', 'decimalLongitude',
'basisOfRecord', 'coordinateUncertaintyInMeters', 'country', 'year', 'datasetKey', 
'institutionCode', 'publishingOrgKey', 'taxonomicStatus' and 'taxonRank'.

\item[\code{time\_period}] Numerical vector. Observations will be downloaded according to the chosen
year range. Default is c(1000,3000). Observations with year = NA are kept by default.

\item[\code{identic\_xy}] Logical. Should records with identical xy be kept?

\item[\code{wConverted\_xy}] Logical. Should incorrectly lon/lat converted xy be kept?
Uses `cd\_ddmm` from 'CoordinateCleaner' R package.

\item[\code{centroids}] Logical. Should species records from raster centroids be kept?
Uses `cd\_round` from 'CoordinateCleaner' R package.

\item[\code{ntries}] Numeric. In case of failure from GBIF server or within the rgbif package, how many
download attempts should the function request? Default is '10' with a 2 seconds interval
between tries. If attempts failed, an empty data.frame is return by default.

\item[\code{error.skip}] Logical. Should the search process continues if ntries failed ?

\item[\code{...}] Additonnal parameters for the function cd\_round() of CoordinateCleaner.
\end{ldescription}
\end{Arguments}
%
\begin{Details}\relax
Argument `grain` used for two distinct gbif records filtering. (1) Records filtering
according to gbif 'coordinateUncertaintyInMeters'; every records uncertainty > grain/2
are removed. Note: Records with no information on coordinate uncertainties are kept by
default. (2) Records filtering according to the number of longitude/latitude decimals;
if 110km < grain <= 11km, lon/lat with >= 1 decimal are kept, if 11km < grain <= 1100m,
lon/lat with >= 2 decimals kept; if 1100m < grain <= 110m, lon/lat with >= 3 decimals
are kept; if 110m < grain <= 11m, lon/lat with >= 4 decimals are kept;
if 11m < grain <= 1.1m, lon/lat with >= 5 decimals are kept etc...
\end{Details}
%
\begin{Value}
Object of class data.frame with requested GBIF information. Although the function
works accurately, error outputs might still occur depending on the 'sp\_name' used.
Therefore, default information detailed in 'add\_infos' is stored so that sanity checks
may still be applied afterwards. Although crucial preliminary checks of species records
are done by the function, addtional post exploration with the CoordinateCleaner R
package is still highly recommended.
\end{Value}
%
\begin{References}\relax
Chauvier, Y., Thuiller, W., Brun, P., Lavergne, S., Descombes, P., Karger, D. N., ... \& Zimmermann,
N. E. (2021). Influence of climate, soil, and land cover on plant species distribution in the
European Alps. Ecological monographs, 91(2), e01433. 10.1002/ecm.1433

Chamberlain, S., Oldoni, D., \& Waller, J. (2022). rgbif: interface to the global biodiversity
information facility API. 10.5281/zenodo.6023735

Zizka, A., Silvestro, D., Andermann, T., Azevedo, J., Duarte Ritter, C., Edler, D., ... \& Antonelli,
A. (2019). CoordinateCleaner: Standardized cleaning of occurrence records from biological collection
databases. Methods in Ecology and Evolution, 10(5), 744-751. 10.1111/2041-210X.13152

Hijmans, Robert J. "terra: Spatial Data Analysis. R Package Version 1.6-7." (2022). Terra - CRAN
\end{References}
%
\begin{SeeAlso}\relax
The (1) rgbif and (2) CoordinateCelaner packages for additional and more general
approaches on (1) downloading GBIF observations and (2) post-filetering those.
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}

# Necessary libraries
#library(raster)
#library(terra)
#library(rgbif)
#library(CoordinateCleaner)

# Load the Alps Extend
data(geo_dat)

# Downloading worldwide the observations of Panthera tigris
test1 = wsl_gbif("Panthera tigris")

# Downloading in the Alps the observations of Cypripedium calceolus (with a 100m grain and
# by adding the 'issues' column)
test3 = wsl_gbif("Cypripedium calceolus", geo = shp.lonlat, grain = 100, add_infos = c("issue"))
plot(shp.lonlat)
points(test1[,c("decimalLongitude","decimalLatitude")],pch=20,col="#238b4550",cex=1)

# Downloading worlwide the observations of Ailuropoda melanoleuca (with a 100km grain, after
# 1990 and by keeping duplicates and by adding the name of the person who collected the species records)
test3 = wsl_gbif("Ailuropoda melanoleuca", grain = 100000 , duplicates = TRUE,
    time_period = c(1990,3000), add_infos = c("recordedBy","issue"))

# Downloading worlwide the observations of Phascolarctos cinereus (with a 1km grain, after 1980,
# and keeping raster centroids)
test4 = wsl_gbif("Phascolarctos cinereus", grain = 1000,
    time_period = c(1990,3000), centroids = TRUE)

\end{ExampleCode}
\end{Examples}
\inputencoding{utf8}
\HeaderA{wsl\_obs\_filter}{Filter a set of GBIF observations according to a defined grain}{wsl.Rul.obs.Rul.filter}
%
\begin{Description}\relax
Filter a set of GBIF observations through a chosen raster grid that defines
the resolution of the spatial analysis. Depending on the desired resolution
of the analysis, one might want to keep only one GBIF observation per grid
cell to avoid e.g., modelling and sampling bias issues.
\end{Description}
%
\begin{Usage}
\begin{verbatim}
wsl_obs_filter(wsl.gbif, grid)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{wsl.gbif}] one wsl.gbif output including one or several species

\item[\code{grid}] Object of class 'SpatRaster', 'RasterLayer', 'RasterBrick' or
'RasterStack' of desired resolution and extent.
\end{ldescription}
\end{Arguments}
%
\begin{Value}
a data frame with two columns named 'x' and 'y' comprising
the new set of observations filtered at grid resolution.
\end{Value}
%
\begin{Examples}
\begin{ExampleCode}

# Load the European Alps extent and a raster of a random resolution
data(geo_dat)
data(exrst)

# Downloading in the European Alps the observations of Arctostaphylos alpinus
obs.arcto = wsl_gbif("Arctostaphylos alpinus",geo=shp.lonlat)
plot(shp.lonlat)
points(obs.arcto[,c("decimalLongitude","decimalLatitude")],pch=20,col="#238b4550",cex=1)

\end{ExampleCode}
\end{Examples}
\inputencoding{utf8}
\HeaderA{wsl\_taXnames}{Retrieve from GBIF all scientific names of a specific Taxa}{wsl.Rul.taXnames}
%
\begin{Description}\relax
Allows to extract from the gbif backbone taxonomy all names from an input species name
(accepted, synonyms, children, related...).
\end{Description}
%
\begin{Usage}
\begin{verbatim}
wsl_taXnames(sp_name = NULL, conf_match = 90, all = FALSE)
\end{verbatim}
\end{Usage}
%
\begin{Arguments}
\begin{ldescription}
\item[\code{sp\_name}] Character. Species name from which the user wants to retrieve all existing GBIF names

\item[\code{conf\_match}] Numeric. From 0 to 100. Determine the confidence
threshold of match of 'sp\_name' with the GBIF backbone taxonomy. Default is 90.

\item[\code{all}] Logical. Default is FALSE. Should all species names be retrieved or only
the accepted name and its synonyms?
\end{ldescription}
\end{Arguments}
%
\begin{Value}
A data.frame with two columns: (1) Names and (2) Backbone Taxonomy Status
\end{Value}
%
\begin{References}\relax
Chauvier, Y., Thuiller, W., Brun, P., Lavergne, S., Descombes, P., Karger, D. N., ... \& Zimmermann,
N. E. (2021). Influence of climate, soil, and land cover on plant species distribution in the
European Alps. Ecological monographs, 91(2), e01433. 10.1002/ecm.1433

Chamberlain, S., Oldoni, D., \& Waller, J. (2022). rgbif: interface to the global biodiversity
information facility API. 10.5281/zenodo.6023735
\end{References}
%
\begin{SeeAlso}\relax
The rgbif package for additional and more general approaches on how to retrieve
scientific names from the GBIF backbone taxonomy.
\end{SeeAlso}
%
\begin{Examples}
\begin{ExampleCode}
wsl_taXnames("Cypripedium calceolus",all=FALSE)
wsl_taXnames("Cypripedium calceolus",all=TRUE)

\end{ExampleCode}
\end{Examples}
\printindex{}
\end{document}
